---
title: "phiX174 Melting Curve Extraction"
author: "Bob"
date: "2025-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Coordinate extraction

Points along the melting curve for the phiX174 bacteriophage were extracted from Figure 3 of Azbel, [DNA sequencing and melting curve](https://pmc.ncbi.nlm.nih.gov/articles/PMC382884/pdf/pnas00001-0109.pdf), published in 1979. Here we fit a spline to these points, and use that to interpolate the melting curve at regularly spaced temperatures.

* solid line: PhiXD, 15 mM NaC/1.5 mM Nacitrate
* dashed line: fragments Y1 and Y2

Here is the sequence of the [PhiX174 bacteriophage](https://www.ncbi.nlm.nih.gov/nuccore/NC_001422.1)
 
[Comparison of theoretical denaturation maps of ϕX174 and SV40 with their gene maps](https://pmc.ncbi.nlm.nih.gov/articles/PMC327754/pdf/nar00444-0261.pdf)

```{r libraries}
library('ggplot2')
```

```{r phiX174}
# setwd("~/Documents/projects/cybertory_TNG/melting_curve_simulation_2025/reference_curves/phiX174_1979")

phiX174_raw <- read.csv("data/solid_line.csv", header=FALSE, col.names=c("T", "dFdT"))

# with (phiX174_raw, plot(temperature, dFdT, type='l'))

fit_ss <- with(phiX174_raw, smooth.spline(T, dFdT, nknots = 60)) # try various numbers of knots 

std_data <- data.frame(T = seq(69.5, 75.0, by=0.1))
std_data['dFdT'] <- predict(fit_ss, std_data['T'])$y

phiX174_raw |>
  ggplot(aes(x=T, y=dFdT)) + 
	geom_line() + 
  geom_point(data=std_data, aes(x=T, y=dFdT), col="red")

std_data |> write.csv("phiX174_std.csv")
```

# To Do:

* predict melting curve for PhiX174
* compare to this experimentl curve.